\documentclass[xcolor=dvipsnames]{beamer}
\mode<presentation>
\synctex=1
\setbeamertemplate{itemize item}[ball]
%\usetheme[secheader]{Boadilla}
%\usetheme{boxes}
\usetheme{myBoadilla}
%\usetheme[secheader]{myBoadilla}
%\usetheme[left]{Goettingen}

%\usepackage{enumitem}

%\usepackage{CSIC2}
\setbeamertemplate{itemize item}[ball]
 \setbeamertemplate{navigation symbols}{
      \insertslidenavigationsymbol
%%      \insertdocnavigationsymbol
%%    {\footnotesize \insertframenumber} %% FIXME: delete if we use CSIC2
 }

\usepackage[absolute,overlay]{textpos}
% \usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage[T1]{fontenc}
% Or whatever. Note that the encoding and the font should match. If T1
% does not look nice, try deleting the line with the fontenc.
%%\usepackage{fancybox}
\usepackage[lined]{algorithm2e}

\usepackage{url}
\newcommand{\cyan}[1]{{\textcolor {cyan} {#1}}}
\newcommand{\blu}[1]{{\textcolor {blue} {#1}}}
\newcommand{\Burl}[1]{\blu{\url{#1}}}
\newcommand{\red}[1]{{\textcolor {red} {#1}}}
\newcommand{\green}[1]{{\textcolor {green} {#1}}}
\newcommand{\mg}[1]{{\textcolor {magenta} {#1}}}
\newcommand{\og}[1]{{\textcolor {PineGreen} {#1}}}
\newcommand{\code}[1]{\texttt{\slshape\footnotesize #1}}
\newcommand{\myverb}[1]{{\footnotesize\texttt {\textbf{#1}}}}
\newcommand{\Rnl}{\ +\qquad\ }
\newcommand{\Emph}[1]{\emph{\mg{#1}}}
\newcommand{\gry}[1]{{\textcolor {gray} {#1}}}

\newcommand{\hanging}[1]{
\renewcommand{\baselinestretch}{0.7}\small\normalsize
\setlength{\parskip}{1.5ex plus0ex minus0ex}
  \noindent\hangindent=24pt
  \hangafter=1
   {\scriptsize #1}
  %\hangindent\parindent
}


\newcommand{\hango}[1]{
\renewcommand{\baselinestretch}{0.7}\small\normalsize
\setlength{\parskip}{1.5ex plus0ex minus0ex}
  \noindent\hangindent=24pt
  \hangafter=1
   {\tiny #1}
  %\hangindent\parindent
}


\newcommand{\hanginp}[2]{
  \renewcommand{\baselinestretch}{0.7}\small\normalsize
  \setlength{\parskip}{1.5ex plus0ex minus0ex}
  \noindent\hangindent=24pt
  \hangafter=1
  {\scriptsize #1}{\hspace{2pt}\Large\red{#2}}
  % \hangindent\parindent
}


%% use for itemize wihtin description
% \newenvironment{myitemize} {
%   \setlength{\leftmarginii}{-10mm}
%   \itemize
% }

% \newenvironment{myitemize}
% {\setlength{\leftmarginii}{-10mm}
% \begin{itemize}}
% {\end{itemize}}


\newenvironment{myitemize}
{\begin{itemize}}{\end{itemize}}

\newenvironment{wideitemize}
{\itemize\addtolength{\itemsep}{10pt}}{\enditemize}

\newcommand{\mydes}[2][]{
  \vspace*{-1pt}
%%  \renewcommand{\baselinestretch}{0.7}\small\normalsize
  \begin{description}\item[{#2}]{#1}\end{description}
\vspace*{-8pt}
}


\usepackage{gitinfo}
\usepackage{wasysym} %% smileys

\title{BM-2: Applied statistics}



\author[R. Diaz-Uriarte]{Ramón Diaz-Uriarte} %% \\ \Burl{http://ligarto.org/rdiaz}}

\institute[]{Department of Biochemistry\\
Universidad Autónoma de Madrid\\
Madrid, Spain\\
\texttt{r.diaz@uam.es}\\
\Burl{https://ligarto.org/rdiaz}
}


% \date[22-09-2014]{22-September-2014}
\date[04-September-2025]{ {\scriptsize (\gitAuthorDate \  Rev: \gitAbbrevHash)}}
% \date[13-09-2019]{13-September-2019\\ {\scriptsize (Rev: \gitAbbrevHash)}}


\begin{document}


\begin{frame}
  \titlepage
\end{frame}


% \begin{frame}
%    \frametitle{Outline}
%  {\small
%  \tableofcontents[subsectionstyle=hide]
%  }
%  \end{frame}



\begin{frame}
  \frametitle{Objective of this intro}
  \begin{itemize}
  \item How and why we will do what we will do
  \item Mechanics of the class
  \item What to expect
    \begin{itemize}
    \item What you can expect
    \item What we expect from you
    \end{itemize}
  \item How to study for it
  \end{itemize}

  %% Tone of this intro somewhat defenssive. Apologies for that. Some slides are here out of frustration with a few repeated issues over the years.
\end{frame}


\begin{frame}
  \frametitle{People}
  \begin{itemize}
  \item Blanca Lizarbe Serra, \texttt{blanca.lizarbe@uam.es}
  \item Ramon Diaz-Uriarte, \texttt{r.diaz@uam.es}
  \end{itemize}
\end{frame}




\begin{frame}
  \frametitle{Why, o why? (Oh, no, statistics again!)}
  \begin{itemize}
  \item (In the past) serious deficiencies in elementary usage of
    statistics have been detected (e.g., during TFM).

\item   Because the world is a noisy place and we need statistics to do science.
  And we always come back to statistics.

  \end{itemize}

% \vspace*{18pt}

% \begin{verse}
%   Go ahead, tell me you'll leave again\\
%   You'll just come back running\\
%   Holding your scarred heart in hand\\
%   It's all the same

%   And I'll take you for who you are\\
%   If you take me for everything\\
%   And do it all over again\\
%   It's all the same
% \end{verse}

% {\fontfamily{ppl}\selectfont Sick Puppies, \textit{All the Same} (from
% their 2007 album \textit{Dressed Up as Life}).}

\end{frame}

\begin{frame}
  \frametitle{Objectives}
  \begin{itemize}
  \item Be able to conduct, and interpret correctly, some \textbf{basic} types of
    statistical analysis of wide use in molecular biology et al.
  \item Develop your statistical intuition.
  \item Understand the fundamental relevance of experimental design.
  \end{itemize}
\end{frame}


\begin{frame}
  \frametitle{Syllabus}
  \begin{itemize}
  \item Reading and handling data. Graphics.
  \item Comparing two groups (and making inferences about a single one).
  \item Comparing more than two groups: ANOVAs and linear models. The multiple testing problem.
  \item Basic categorical data analysis.
% \vspace*{0.5cm}
% \item All sessions: computer work
  \end{itemize}
\end{frame}



\begin{frame}
  \frametitle{Why those topics}
  \begin{itemize}
  \item Indispensable
  \end{itemize}
\end{frame}



\begin{frame}
  \frametitle{Managing expectations}
  Is this all the stats knowledge I need for \ldots

\vspace*{15pt}
  \begin{itemize}
  \item My ``trabajo fin de master''? Most likely not.
  \item Doing research for the rest of my life? \textbf{Definitely not}.
  \end{itemize}
\end{frame}



\begin{frame}
  \frametitle{\ldots eh????!!!!}
  \begin{itemize}
  \item We cannot cover all you might need
  \item Develop statistical intuition: ease further learning
  \item Know when to consult a statistician ({\footnotesize hint: as soon
      as possible})
  \end{itemize}
\vspace*{10pt}

What are we leaving out? Almost everything.


\vspace*{10pt}


{\footnotesize
(Analogy of first aid/CPR course vs.\ becoming a neurosurgeon)
}

\end{frame}




\begin{frame}
  \frametitle{Statistics (with R)}
  \begin{itemize}
  \item Key concepts:
    \begin{itemize}
    \item basic statistical analysis
    \item basic graphical procedures
    \end{itemize}
    \vspace*{10pt}
  \item But this is done with something (some piece of software).
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Why R?}
  \begin{itemize}
  \item Free (as in ``free beer'' and as in ``free speech''): free software
  \item Available for all major operating systems
  \item High quality
  \item Huge library of contributed packages with added
    functionality (including BioConductor); increasing weekly
  \item The most used statistical system in bioinformatics and
    ``omics'' research
  \item Available in popular ``scientific workflows''
    (\Burl{http://en.wikipedia.org/wiki/Scientific_workflow_system}) such
    as Kepler, Taverna, Galaxy or Knime.
  \item A key piece in reproducible research
  \item They even started to mention it in \textit{Nature} editorials.
    % (\textit{The digital toolbox}, 4-September-2014) and \textit{Nature}
    % toolboxes articles (Tippman, 2015, \textit{Programming tools:
    %   adventurs with R}, January 2015) :-) This is now common nowadays.
  \item \ldots

  \end{itemize}
\end{frame}








\begin{frame}
  \frametitle{How? R Commander}
  \begin{itemize}
  \item We will use R Commander (\Burl{https://cran.r-project.org/web/packages/Rcmdr/index.html})
  \item GUI + full transcript + markdown
  \item Also RStudio (secondary)
  \item But you will also have to type commands
  \end{itemize}
\end{frame}


% \begin{frame}
%   \frametitle{How does it look?}
% \vspace*{4cm}
% {\footnotesize(I'll show here a couple of desktops with RStudio and R Commander)}
% \end{frame}



%%% Say: I do not use for real, ever.


% \begin{frame}
% %%  \frametitle{(Don't tell anyone)}
%   \frametitle{Our hidden agenda}
%   \begin{itemize}
%   \item That you finish the class with a burning desire (to learn) to use
%     R directly typing R commands (i.e., no mouse clicks).
%   \item That you finish the class with a burning desire to learn more
%     statistics.
%   \end{itemize}
% \end{frame}



\begin{frame}
  \frametitle{And what if I know R already?}
  \begin{itemize}
  \item Great!
  \item Focus on the statistics (we provide the commands that R Commander generates).
  \item You can ignore R Commander almost completely if you want.
  \item Again: emphasis is on statistics, not R. ({\tiny You can go through
    the course without using R even once.})
  \end{itemize}
\end{frame}



\begin{frame}
  \frametitle{What should  happen after we are done?}
 (For example \ldots)
  \begin{itemize}
  \item That you never again use Excel for figures.
  \item That you do your statistics with R, from simple t-tests to complex
    GLMs.
  \item That you will continue  to practice (otherwise, you'll forget the material).
  \end{itemize}
\end{frame}


\begin{frame}
  \frametitle{What do I need?}
  \begin{itemize}
    \item A computer with R. %%You can bring your own laptop to class.
      \begin{itemize}
      %% \item But you do not need to.
      \item During classes themselves little time devoted to using the
        computer. (More about this later).
      \end{itemize}
\vspace*{15pt}
%%\item Computers in class will have software installed
  \item You \textbf{\red{MUST}} install several things on your own computer
    \begin{itemize}
    \item for the course %if you bring your computer to class
    % \item for out-of-class work
    \item for the rest of your life
    \end{itemize}
  \end{itemize}
\end{frame}


%   \textbf{STOP! Some packages available in July and August 2014 are now
%     deprecated. On 19-September-2014 I will provide updated notes that
%     will work as of that day}.

% \textbf{Thus, the text below is mostly, but not fully, correct as of now.}

%   See   \Burl{http://ligarto.org/rdiaz/BM1-software-applied-stats.txt}
% \vspace*{10pt}

% or


\begin{frame}
  \frametitle{Software in your laptops}

  Instructions available from Moodle: file \textbf{``Software-you-need-to-install.txt''}
  % and also from
  % \Burl{https://github.com/rdiaz02/BM-1/blob/master/2022-2023/BM2-software-applied-stats.txt}.
  %   \vspace*{12pt}

(If it helps, there is also a video of the process in Moodle.)


  % \vspace*{12pt}
  % {\small(Google: ``github rdiaz bm1'')}


  \vspace*{20pt}

  {\LARGE \textbf{Please, please, please: follow instructions carefully.}}

  \vspace*{15pt}

  \textbf{Read software installation instructions at least once BEFORE starting the installation process.}

\end{frame}





% \begin{frame}
%   \frametitle{Two more packages: RcmdrPlugin.doBy and ISwR}
% We need to add a couple of things.
%   \begin{itemize}
%   \item You also will want to install the package ``ISwR'': I use one data
%     set for a couple of examples.
%     %% FIXME!!!
% \item Download it from \red{zz FIXME}:
%   \begin{itemize}
%   \item Linux and Mac:
%   \item Windows:
%   \end{itemize}

% \item You can try installing it. We will also explain it in class.
% \item But you MUST have downloaded it.
%   \end{itemize}
% \end{frame}

% \begin{frame}
%   \frametitle{One more package: RcmdrPlugin.doBy}

%   \begin{itemize}
%   \item You also need to install a modified version of the package ``RcmdrPlugin.doBy''.
%     %% FIXME!!!
% \item Download it from Moodle (file ``RcmdrPlugin.doBy'')

% \item You can try installing it. We will also explain it in class.
%   \begin{itemize}
%   \item In RStudio go to ``Tools'', ``Install Packages''.
%   \item In the entry ``Install from'', select ``Package Archive File''
%     \item In the second entry, ``Package archive'', select the file you
%       got (RcmdrPlugin.doBy\_0.1-4.tar.gz)
%     \item Install
%     \end{itemize}
% \item But you MUST have downloaded it.
%   \end{itemize}
% \end{frame}


\begin{frame}
  \frametitle{Software in your laptops}
  \textbf{\red{Please make sure things work BEFORE coming to class}}
\vspace*{15pt}


And you will need the software to work so that you can work on the class
notes \textbf{before} class.

\end{frame}



% \begin{frame}
%   \frametitle{Software in your laptops: solving issues}
%   \begin{itemize}
%   \item 18-September-2019: we will go over   issues in software
%     installation. (12:00 to 14:00, Aula 7).
%   \item \textbf{Please, come after having tried on your own!!!}
%     \begin{itemize}
%     \item not enough bandwith
%     \item not enough time
%     \end{itemize}
%   \item {\small Come only if you had trouble with installation. No need to
%     come otherwise.}
%   \end{itemize}
% \end{frame}



\begin{frame}
  \frametitle{Can I use \ldots}
  \begin{itemize}
  \item SAS
  \item SPSS
  \item Minitab
  \item Matlab
  \item Stata
  \item Statgraphics
  \item Excel (not this one, unless you use Rexcel or similar!)
  \item \ldots
  \end{itemize}
  yes, of course!! \ldots
  \pause

\vspace*{10pt}

\ldots but you are on your own: we will not provide any help for systems
  other than  R {\scriptsize (the ``official'' R from CRAN)} with version $\ge$ 4.5.1.
\end{frame}





\begin{frame}
   \frametitle{But the objective of this course is not to teach you R}
   \begin{itemize}\addtolength{\itemsep}{3.1ex}
   %% \item \ldots that would be too ambitious
   \item \ldots the material in the notes will allow you to learn quite
     a bit of R.
   \item Aren't the last slides inconsistent?
   \item Nope:
     \begin{itemize}
     \item Course objective: learn basic stats.
     \item We need a common system (``communicate in a shared language when
       using the computer'') to do stats: that will be R. This is the
       vehicle.
     \item (Other analogies come to mind, like teaching the class in
       Spanish, English, Portuguese, French, German, Esperanto, \ldots)
     \end{itemize}
   \end{itemize}
\end{frame}




\begin{frame}
  \frametitle{R Commander, R Studio, and R are NOT the objective of the class}
  R Commander, R Studio, and R will NOT be things we ask about in the exam!!
  \begin{itemize}
  \item We need a common language, a common vehicle to teach stats
  \item The purposes of the course is to learn stats, not R or R Studio or R Commnader
  \item In the exam you will NOT use a computer
  \item In the exam you will NOT be asked questions about R or R Commander or R Studio. But you must be able to interpret output, including graphics and output from statistical analyses.
  \item \textbf{Any questions about this?}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Schedule}
  \begin{itemize}
  \item From Moodle. \textbf{Schedule-applied-statistics-2025-26.pdf}
\vspace*{10pt}
  \item Two different groups:

      \begin{itemize}
      \item BT and BDC: Blanca Lizarbe
      \item BMM: Ramon Diaz-Uriarte
      \end{itemize}
  % \item Pay attention to days 23, 24, 25, 28 September and groups A and B (B
  %   has class on 24 and 25 for Lesson 1, A has class on 23 and 28 for
  %   Lesson 1).

      \vspace*{10pt}
  \item Exam: same day for everybody. Two groups, same two groups as for classes.
  \end{itemize}
\end{frame}



\begin{frame}
  \frametitle{Two groups}
  \begin{itemize}\addtolength{\itemsep}{3.1ex}
  \item Each group has a different Moodle: different forum for questions, possibly some slightly different material, etc, etc. (Not yet!! See later)
  \item Each group might have a slightly different exam.
  \item \textbf{Please, go to the class you are assigned to}
  \end{itemize}
\end{frame}



\begin{frame}
  \frametitle{Schedule: let us look at it}
  (Ramon is supposed to open the PDF with the schedule)
\end{frame}


\begin{frame}
  \frametitle{Class mechanics}
  (Now we show a bunch of slides, but all of them say basically the same thing :-) : Work on the notes BEFORE coming to class. )

\end{frame}

\begin{frame}
  \frametitle{Class mechanics}
  \begin{itemize}\addtolength{\itemsep}{3.1ex}
    %% \item A mixture of theory and practice
  \item \textbf{You must work through the notes before coming to class}
  \item Introduce a problem, work through it
  \item Explain the relevant elements of theory, as needed
  \item You need to work through the notes before class
  \item Yes, there will be partial repetition between
    classes/sections
  \end{itemize}
\end{frame}




\begin{frame}
  \frametitle{Class mechanics}
  \begin{itemize}\addtolength{\itemsep}{3.1ex}
  \item Do NOT expect too much hand holding in terms of  ``where is the
    pull down menu for importing a file'', etc
  \item We (teachers) will:
    \begin{itemize}
    \item Explain concepts that you do not understand
    \item Help you get going
    \item Try to save you from despair
    \end{itemize}
  \item Learning R is not the objective
  \item You do the statistical learning work: learn by
    doing/exploring. \textbf{Active learning}
  \end{itemize}
\end{frame}






\begin{frame}
  \frametitle{Working through the notes before class}
  Why?


\vspace*{15pt}
Because you will learn the material a lot better (\textbf{Active
  learning})



\begin{itemize}\addtolength{\itemsep}{3.1ex}
\item try it on your own
\item find out what you don't know/understand (or makes no sense in the notes)
\item get it answered (you get a need fulfilled)
% \item \textbf{Even more important with on-line teaching} (Trying to follow
%   a 2-hour stats class via Teams just passively listening \ldots not a
%   good experience.)
\end{itemize}
\end{frame}


\begin{frame}
  \frametitle{Working through the notes before class}
  Why?


\vspace*{15pt}
Because there is a lot of material to cover (really: A LOT ---look at the
number of pages and topics in the notes)

\begin{itemize}\addtolength{\itemsep}{3.1ex}
\item No time to cover every single thing in the notes
\item If we get the trivial stuff out of the way, we can deal with more
  interesting stuff
%% \item And again: on-line teaching.
\end{itemize}

\end{frame}



\begin{frame}
  \frametitle{Working through the notes before class}
  Can you really do it? \textbf{YES!}

\vspace*{15pt}

Because all of you should have taken at least one statistics class so you can look at your old stats class notes (and
for specific things you need to remember, there is always Wikipedia).

\vspace*{15pt}


\end{frame}


% \begin{frame}
%   \frametitle{Working through the notes before class}

% Thus \ldots

% \begin{itemize}
% \item The Applied Statistics part of BM-1 should be a quick review \ldots
% \item focused on things you might have forgotten or tricky things you
%   might not know \ldots
% \item and with a strong practical component of ``teach me how to do it
%   with available software, interpret it, and produce reports and graphics
%   I can use in my TFM (and life at large)''.
% \end{itemize}

% \end{frame}





\begin{frame}
  \frametitle{Working through the notes before class}

How? Up to you.

\vspace*{15pt}

Working in groups a good idea

\begin{itemize}
\item Your peers are likely to be the ones who best understand what it is you do not
  understand
\end{itemize}

\end{frame}




% \begin{frame}
%   \frametitle{Working through the notes before class}
%   Please, please: do it
% \vspace*{15pt}
%   \begin{itemize}
%   \item
%   \end{itemize}
% \end{frame}


\begin{frame}
  \frametitle{Working through the notes before class}
%  Please, please: do it
  \begin{itemize}\addtolength{\itemsep}{3.1ex}
  \item If none of you work on this on your own \ldots
    \begin{itemize}
    \item classes will go painfully slow
    \item we will all end up frustrated
    \end{itemize}
  \item If some of you work through it and some won't \ldots most
    likely those who don't will be left way behind ---and ``tutorías'' are
    not the way to fix not working through the notes before class
  \end{itemize}
\end{frame}


\begin{frame}
  \frametitle{In case it was not clear}
  Please,  work on the notes on your own before coming to class.

\vspace*{15pt}


What you get out of this class depends a lot on what you put into
    this class
\end{frame}



\begin{frame}
  \frametitle{Class notes}
  \begin{itemize}\addtolength{\itemsep}{3.1ex}
  \item Available before class
  \item If you can, avoid printing them:
    \begin{itemize}
    % \item all major operating systems have
    % PDF viewers that allow you to incorporate annotations in the PDF
    % itself and save them for posterity;
  \item notes might suffer minor changes up until right before class;
  \item ($\Rightarrow$ if you are reading these notes before class,
    these notes themselves can suffer changes :-) )
    \end{itemize}
\end{itemize}
\end{frame}



\begin{frame}
  \frametitle{Class notes and following the order}
  \begin{itemize}\addtolength{\itemsep}{3.1ex}
  \item There might be slight deviations from the sequential order of pages in the notes. % Do not expect as to follow the notes being faithful to every single comma or period.

  \item There are topics in the notes that we will not cover. This is on purpose, and the notes anticipate it (e.g., ``This is advanced material and we will not cover it in class'').
  \item We will use additional material in other files, e.g., other PDFs (and this is indicated from the main notes).
  \item We might, if time permits, add additional examples, or small exercises to work at home and discuss in class (this will not be graded).
  \item \textbf{Questions about class mechanics?}
  \end{itemize}
\end{frame}



\begin{frame}
  \frametitle{Tutorías}
  \begin{itemize}\addtolength{\itemsep}{3.1ex}
  \item Tutorías are there for a reason
  \item But tutorías are \textbf{not}:
    \begin{itemize}
    \item Remedial classes if, for some reason, you cannot come to a class, \textbf{except for exceptional, well justified reasons}. If you cannot come to class, ask your classmates and work on your own with the class notes.
    \item Remedial introductory sessions (i.e., do not use a tutoría if you could not come here today).
    \item Remedial ``R installation Problems'' sessions: software installation \textbf{must} be attempted on your own before 11-September. If you have doubts or difficulties, ask in the forum and come to class on 11-September.
    \end{itemize}
  \end{itemize}
\end{frame}


\begin{frame}
  \frametitle{Applied Statistics Forum}
  \begin{itemize}\addtolength{\itemsep}{3.1ex}
  \item Use the forum for questions.
  \item Do not expect us to always be the ones who answer.
  \item \textbf{Label your questions appropriately}. The title/subject of
    the question is \textbf{very} important.
    \begin{itemize}
    \item Searching for answers to recurrent questions.
    \item Knowing who should answer the question.
    \end{itemize}
  \end{itemize}
\end{frame}


\begin{frame}
  \frametitle{Forum: examples of poor question titles/subjects}

  \begin{wideitemize}\addtolength{\itemsep}{3.1ex}
  \item R does not run. \textit{What does not run, exactly?}
  \item I cannot install R. \textit{Where and what fails?}
  \item I don't understand statistics. \textit{Eh?! None of
      statistics? }
  \item I hate statistics. \textit{Fair enough. Now, what exactly is the
      question? Or is this just to make us sad?}
  \item I do not understand a t-test. \textit{What don't you understand?
      The formula? The null hypothesis? The \ldots?}
  \end{wideitemize}
\end{frame}



\begin{frame}
  \frametitle{Forum: examples of good question titles/subjects}
  \begin{wideitemize}
  \item Lesson 3, Section 10.2, Diagnostics regression: why outliers?
  \item Installing R: cannot install rJava.
  \end{wideitemize}
\end{frame}



\begin{frame}
  \frametitle{Forum: attachments and error messages}
  \begin{itemize}\addtolength{\itemsep}{3.1ex}
  \item Whenever possible, copy and paste the error message(s) as
    \textbf{text} in the body of the question.

  \item Avoid taking a screen capture and pasting it in the email message.
    \item Avoid taking a screen capture  and pasting it in a word document, power  point document, or other such perverse thing.

  \item Text is easy to search. A png isn't. A doc and a pdf aren't as
    bad as an image, but not as good.

  \item Text is easy to copy/modify.

  \item (Yes, copying and pasting the text error in an email message is
    much simpler and faster for you too.)
  \end{itemize}
\end{frame}




\begin{frame}
  \frametitle{Moodle}
  %\begin{itemize}\addtolength{\itemsep}{3.1ex}
  %\item A few technical issues that need to be solved
  %\item As of today, a single one for everybody. It only has the bare essentials:
    \begin{itemize}\addtolength{\itemsep}{3.1ex}
    \item Schedule
    \item Software you need to install
    \item A video of software installation
    \item This presentation
    \item Guía docente
    \end{itemize}
  %   \item Soon, two different moodles, one per group: BDC + BT vs. BMM.
  %     \begin{itemize}
  %     \item How will you know? It will either be obvious and/or we will tell you. No need to worry. \smiley{}
  %     \end{itemize}
  % \end{itemize}
\end{frame}


\begin{frame}
  \frametitle{The first key lesson}
\end{frame}


\begin{frame}
  \frametitle{The first key lesson}
  \begin{itemize}\addtolength{\itemsep}{3.1ex}
  \item Both Blanca Lizarbe and myself have \textbf{failed the mind reading
    courses}.
  \item So ...
    \begin{itemize}
    \item ask whatever is not clear
    \item let us know if the pace is too slow
    \item let us know if the pace is too fast
    \end{itemize}
  \end{itemize}
\end{frame}



\begin{frame}
  \frametitle{Grading}
  \begin{itemize}\addtolength{\itemsep}{3.1ex}
  \item Look at guia docente.
    \begin{itemize}
    \item Available from Moodle (and from % \Burl{https://secretaria-virtual.uam.es/doa/consultaPublica/look[conpub]VolverInfoPubGuiaDocAs})

    % \item Also from
\Burl{https://secretaria-virtual.uam.es/doa/consultaPublica/look[conpub]BuscarPubGuiaDocAs?entradaPublica=true&idiomaPais=es.ES&_centro=106&_planEstudio=764} and from the ``Secretaría virtual UAM'' ---google for ``guias docentes UAM master'')

% \Burl{https://secretaria-virtual.uam.es/doa/consultaPublica/look[conpub]BuscarPubGuiaDocAs?entradaPublica=true&idiomaPais=es.ES&_anoAcademico=2021&_centro=106&_planEstudio=764}

      \end{itemize}
  \item A final exam with practical cases (70\% of the grade).
  % \item (We need to see how wifi/network access works for real to decide
  %   if in-class or out-of-class)
  \item Self-evaluation exercise, similar to the exam, that you can do on
    your own during the course (30\% of the grade).
    \begin{itemize}
\item \textbf{Please, pay attention to the deadlines of the self-evaluation!!!! Once closed, it is closed.}
    \end{itemize}
    %% \item To pass: at least a 5 over 10 in the final exam.
  \end{itemize}
\end{frame}




\begin{frame}
  \frametitle{And where/when will classes be held?}
  \begin{itemize}\addtolength{\itemsep}{3.1ex}
  \item Look at the schedule.
  \item Always at Facultad de Medicina except the exam (Ciencias).
  % \item All classes on-line, synchronous, via Teams (``presencial
  %   on-line'') or on-site.
  % \item Exam: on-site (presencial físico). (Unless things really change for the worse).
  \end{itemize}
\end{frame}



\begin{frame}
  \frametitle{Schedule: let's see it again}
  As it says

\vspace*{12pt}

  First day: only if you have trouble with installation.
\end{frame}



% \begin{frame}
%   \frametitle{Using my own computer}
%   \begin{itemize}
%   \item You are very welcome to bring your own laptop to class.
%   % \item Make sure you install the required software %% {\scriptsize  \Burl{http://ligarto.org/rdiaz/BM1-software-applied-stats.txt}}
%   % \item \textbf{Wi-Fi (eduroam)}: (google for ``wifi UAM'') %\Burl{http://www.uam.es/servicios/ti/servicios/wifi/}
%   % \item \textbf{\red{Please make sure it works BEFORE coming to class}}
%   % \item Address Wi-Fi questions/issues to CAU (we cannot help).
%   \item Note: Wi-Fi/internet should not be really needed for classes, but it
%     will be needed for the exam.

%     \item For the exam you will need either a laptop with Wi-Fi, or a
%       smartphone or table with an internet connection: you will do the
%       exam in Moodle.
%   \end{itemize}
% \end{frame}

\begin{frame}
  \frametitle{Exam: using my own computer}
  \begin{itemize}\addtolength{\itemsep}{1.5ex}
  % \item You are very welcome to bring your own laptop to class.
  % % \item Make sure you install the required software %% {\scriptsize  \Burl{http://ligarto.org/rdiaz/BM1-software-applied-stats.txt}}
  % % \item \textbf{Wi-Fi (eduroam)}: (google for ``wifi UAM'') %\Burl{http://www.uam.es/servicios/ti/servicios/wifi/}
  % % \item \textbf{\red{Please make sure it works BEFORE coming to class}}
  % % \item Address Wi-Fi questions/issues to CAU (we cannot help).
  % \item Note: Wi-Fi/internet should not be really needed for classes, but it
  %   will be needed for the exam.

  % \item Note: Wi-Fi/internet will be needed for the exam.

    \item For the exam you will need either a laptop with Wi-Fi, or a
      smartphone or tablet with an internet connection: you will do the
      exam in Moodle.
    \item Try \textbf{not} to use a smartphone.
    \item \textbf{Make sure you know how to access Moodle. For example, your password}.
    \item \textbf{Make sure you have a device you can use to take the exam}.
    \item \textbf{Make that device can connect to Moodle, i.e., you use UAM's Eduroam, or your phone as hotspot, or \ldots}
    \item \red{\textbf{\Large{MAKE SURE ALL OF THE ABOVE ARE WORKING BEFORE THE EXAM}}}
  \end{itemize}
\end{frame}



\begin{frame}
  \frametitle{}
  \centerline {\Large Questions?}
\end{frame}


% \begin{frame}
%   \frametitle{}
%   \begin{itemize}
%   \item Who is planning on bringing his/her own laptop?
%   \end{itemize}
% \end{frame}


% 1.  Average mark on Self-evaluation quiz (the post-class quiz). A skipped quiz
% will count as a zero value in order to calculate the average. Please note that
% answers to Problem Quiz (the pre-class quiz) do NOT contribute to this average
% value.

% 2.  Grade on the Final exam. Taking this exam is not mandatory (but it is highly
% recommended).

% The final mark for bioinformatics will be calculated as:

% (0.6xSelf-Eval. average)+(0.4xExam)


\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
